The Ranked Nearest Neighbour Interchange space

### 2019-06-13 paper.tex
restructure + minor edits on caterpillar convexity

@@

@@

@@

    Let $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu)$.
    We can compute a  shortest caterpillar path from $T$ to $T'$ using Bubble Sort.
    In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-2-k$ exchanges of taxa.
    It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,T') = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,T') = \Delta(\rnniu)$.

@@

    Induction step: $n \to n+1$\\
    Induction hypothesis: For trees $T$, $T'$ on $n$ taxa it is $d_c(T,T') = d(T, T') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
    Let $T, T'$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu_{n+1})$.
    Since we can reach $T'$ from $T$ by using Bubble Sort where each pair of taxa that is compared in that algorithm exchanges, we can conclude that taxon $n+1$ must be in the cherry of $T'$.
    It follows that there are on every path (not only caterpillar paths) from $T$ to $T'$ $n-1$ $\rnni$ moves needed to move the parent of $n+1$ down in the tree, since each $\rnni$ move can decrease the rank of an internal node by at most one.
    Let us consider the trees $T|_n$ to $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$.
    When running Bubble Sort to compute a path from $T|_n$ to $T'|_n$ one can observe the same exchanges of taxa as on the caterpillar path from $T$ to $T'$, except for the $n-1$ ones that involve taxon $n+1$.
    Hence, the caterpillar distance is $d_c(T|_n, T'|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T|_n$ and $T'|_n$ and know that $d(T|_n,T'|_n) = \frac{(n-1)(n-2)}{2}$.

    In general, deleting taxon $n+1$ and its parent in every tree on any path $p$ from $T$ to $T'$ results in a path $p'$ from $T|_n$ to $T'|n$.
    Such a path $p'$ from $T|_n$ to $T'|n$ does not contain moves involving the parent of $n+1$.
    Since the parent of $n+1$ need to move from the root of $T_1$ to the cherry of $T_2$ and each $\rnni$ move can change the rank of a node by at most one, $p'$ is at least $n-1$ $\rnni$ moves shorter than $p$. \todo{Really?}
    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, we could delete taxon $n+1$ and at least $n-1$ moves of it's parent and receive a path from $T|_n$ to $T'|_n$ of length less than $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So it would be $d(T|_n, T'|_n) < \Delta(\rnniu_n)$.
    Since this is a contradiction to $d(T|_n, T'|_n) = \Delta(\rnniu_n)$, it must be $d(T,T') = \Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2} $.

@@

%This lemma might be central  for induction proofs in RNNI space
%Is it correct and if it is, do we need to prove the statement, or is it rather obvious?
%The lemma implies another algorithm to compute distances (it would not give shortest paths explicitly, unlike alg:find_path)
\begin{lemma}
    Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $j$ in tree $T_i$.
\end{lemma}

>>>>

%ranks of internal node shall be from 1 to n, leaves have rank 0

@@

%convexity (there is a shortest path within that set, not: all shortest path are within this set)

@@

%This lemma might be central  for induction proofs in RNNI space
%Is it correct and if it is, do we need to prove the statement, or is it rather obvious?
%The lemma implies another algorithm to compute distances (it would not give shortest paths explicitly, unlike alg:find_path)
\begin{lemma}
    Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $j$ in tree $T_i$.
    \label{lemma:distance_delete_taxon}
\end{lemma}

@@

    Let $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu)$.
    We can compute a  shortest caterpillar path from $T$ to $T'$ using Bubble Sort.
    In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-1-k$ exchanges of taxa.
    It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,T') = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,T') = \Delta(\rnniu)$.

@@

    Induction step: $n \to n+1$\\
    Induction hypothesis: For trees $R$, $R'$ on $n$ taxa it is $d_c(R,R') = d(R, R') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
    Let $T, T'$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu_{n+1})$.
    When applying the Bubble Sort algorithm to compute a path between $T'$ and $T$, each pair of taxa that is compared in that algorithm exchanges, since there must be $\Delta(\rnniu)=\frac{(n-1)(n-2)}{2}$ moves between $T$ and $T'$.
    So we can conclude that taxon $n+1$ must be in the cherry of $T'$.
    Let us consider the trees $T|_n$ to $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$.
    When running Bubble Sort to compute a path from $T|_n$ to $T'|_n$ one can observe the same exchanges of taxa as on the Bubble Sort path from $T$ to $T'$, except for the $n-1$ ones that involve taxon $n+1$.
    Hence, the caterpillar distance is $d_c(T|_n, T'|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T|_n$ and $T'|_n$ and know that $d(T|_n,T'|_n) = \frac{(n-1)(n-2)}{2}$.

    With Lemma~\ref{lemma:distance_delete_taxon} we know that the distance between $T|_n$ and $T'|_n$ is at least $d(T|_n, T'|_n) \leq d(T,T') - |r_{T}(p_{n+1}) - r_{T'}(p_{n+1})|$.
    We also know $|r_{T}(p_{n+1}) - r_{T'}(p_{n+1})| = n-1$
    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, it would follow that $d(T|_n, T'|_n) \leq d(T,T') - |r_{T}(p_{n+1}) - r_{T'}(p_{n+1})| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    So it would be $d(T|_n, T'|_n) < \Delta(\rnniu_n)$.
    Since this is a contradiction to $d(T|_n, T'|_n) = \Delta(\rnniu_n)$, it must be $d(T,T') = \Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2} $.

@@
